---
title: "ECON457 R lab 2"
subtitle: "Optimization in R"
author: "Jon Duan | University of Victoria"
date: "ECON403 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
knitr::opts_chunk$set(
  fig.align="center", fig.width=6, fig.height=4.5, 
   out.width="748px", out.length="520.75px",
  dpi=300, fig.path='Figs/',
  cache=T, warning=F, message=F #, echo=F,
  )
```

# Table of contents

1. [Prologue](#prologue)

2. [Optimization basics](#basics)

3. [Linear programming](#linear)

4. [Quadratic programming](#quadratic)

5. [Summary](#summary)


---
class:  inverse, center, middle
name: prologue

# Prologue

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>


---

## Bridge-in


- Optimization problems are everywhere
  - An investor decides how to put her/his money in different assets.
  - A farmer decides how to use her/his land to grow 

---
# Intended Learning Outcome

- Install packages in RStudio
- Recognize the types of optimization problems
- Convert linear programming and quadratic programming to standard forms  
- Use ROI to solve linear programming and quadratic programming problem.
 


---
class: inverse, center, middle
name: basics

# Optimization basics

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>


---
# Information about optimization in R

- [Optimization](http://cran.r-project.org/web/views/Optimization.html)

- [ROI](https://www.rdocumentation.org/packages/ROI/versions/0.2-1)

This file can be downloaded at https://github.com/jonduan/ECON403/blob/master/econ403-lab02.Rmd

or visit https://bit.ly/2C6I5l8


---
# The R Optimization Infrastructure (ROI)

The R Optimization Infrastructure ([ROI](https://cran.r-project.org/web/packages/ROI/ROI.pdf)) package promotes the development and use of open source optimization problem solvers for R.


ROI handle LP up to MILP and MIQCP problems using the following supported solvers :

1. lpSolve

2. quadprog

3. Rcplex

4. Rglpk (default)

5. Rsymphony

more... 

---
#  Solvers

[CPLEX](https://www.r-bloggers.com/using-cplex-in-r-installing-cplexapi-in-windows-10/) and [Gurobi](https://cran.r-project.org/web/packages/prioritizr/vignettes/gurobi_installation.html) are commercial solvers. Students can get academic license. 

[GLPK](http://winglpk.sourceforge.net/) is free. 

R has the high level interface [Rglpk](https://cran.r-project.org/web/packages/Rglpk/index.html).





---
# Pre-assessment

- Linear programming
- Quadratic programming
- R packages 


---
# Install all the packages.

Uncomment the code in follow chunk. I

Install and load all packages.

```{r}
#install.packages("ROI")
#install.packages(c( "Rglpk","ROI.plugin.symphony","ROI.plugin.glpk","ROI.plugin.quadprog","ROI.plugin.nloptr","ROI.plugin.ipop","ROI.plugin.ecos"))
# suppressMessages(library(lpSolve))
# suppressMessages(library(lpSolveAPI))
# suppressMessages(library(quadprog))
# suppressMessages(library(Rglpk))
suppressMessages(library("ROI"))
ROI_registered_solvers()
```


---
class: inverse, center, middle
name: linear

# Linear Progarmming

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---
# Standard form for linear programming

Objective function:
$$f(x) =  c^T x.$$ 
Here, $x$ is a vector in $\mathbb{R}^n$, $c$ is a constant vector in $\mathbb{R}^n$.

A system of linear constraints on the vector $x \in \mathbb{R}^n$.

$$ Lx <= b $$ 

One example:


$$\begin{align*}     \text{maximize: }   2 x_1 &+ 4 x_2 + 3 x_3\\ \text{subject to: }     3 x_1 &+ 4 x_2 + 2 x_3 & <= 60 \\                        2 x_1 &+   x_2 +   x_3 & <= 40 \\                           x_1 &+ 3 x_2 + 2 x_3 & <= 80 
\end{align*}$$



---

# Standard form to ROI package


```{r}
## Simple linear program.
## three decision variables: x1, x2, x3
## matrix form:    objective: c*x 
##               constraints: L*x <= b  
## maximize:   2 x_1 + 4 x_2 + 3 x_3
## subject to: 3 x_1 + 4 x_2 + 2 x_3 <= 60
##             2 x_1 +   x_2 +   x_3 <= 40
##               x_1 + 3 x_2 + 2 x_3 <= 80
##               x_1, x_2, x_3 are non-negative real numbers

LP <- OP( c(2, 4, 3), # objective function, vector c 
          L_constraint(L = matrix(c(3, 2, 1, 4, 1, 3, 2, 1, 2), nrow = 3), # linear constraint: Maxtrix A, r is organized by column not row. 
                       dir = c("<=", "<=", "<="), # direction
                       rhs = c(60, 40, 80)),      # right hand side: vector b 
          max = TRUE )  # defaults is minimize: max = FALSE. change to max
LP # model named as LP
```





---
# Solve

```{r}
sol <- ROI_solve(LP)# , solver = "glpk")  # Solve model LP with specific solver glpk or automatically choose by ROI.
sol
```



---
# Solutions

```{r}
solution(sol, type = c("primal")) # solution for primal problem
solution(sol, type = c("dual"))  # solution for dual problem: minimize total cost by subject to L'y >= c
```

---
# Extract solutions
```{r}
sol$solution
# sol$objval
sol$status
sol$message # detailed message for solution 
```


---
# Hands on a small problem

[Example:](https://www.analyticsvidhya.com/blog/2017/02/lintroductory-guide-on-linear-programming-explained-in-simple-english/) 

- Consider a chocolate manufacturing company which produces only two types of chocolate – A and B. 

- Both the chocolates require Milk and Choco only.  To manufacture each unit of A and B, following quantities are required:

  - Each unit of A requires 1 unit of Milk and 3 units of Choco
  - Each unit of B requires 1 unit of Milk and 2 units of Choco

- The company kitchen has a total of 5 units of Milk and 12 units of Choco. 

- On each sale, the company makes a profit of

  - Rs 6 per unit A sold
  - Rs 5 per unit B sold.


---
class: inverse, center, middle
name: quadratic

# Quadratic Programming

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---
# Quadratic Programming

we can write the quadratic program (QP) compactly as: 

$$\left\{ \begin{array}{l} \mathrm{minimize}_{x \in \mathbb{R}^n}: \qquad F(x) = \frac{1}{2} x^T Q x - L_1^T x  \\ \mathrm{subject\; to:} \qquad L_2x <= b \end{array} \right.$$

Here, $x$ is a vector in $\mathbb{R}^n$, $Q$ is an $n \times n$ symmetric positive definite matrix, $L_1$ is a constant vector in $\mathbb{R}^n$. The function $Q(x)$ is sometimes referred to as a quadratic form and it generalizes the quadratic function $q(x) = ax^2 +bx + c$ to higher dimensions. 




---
# Standard form for quadratic programming problem

It will require putting problem in matrix form.

# Example #1:
Consider the objective function: 

$$\begin{eqnarray*} F(x,y) &=& \frac{1}{2} \left[ \begin{matrix} x & y \end{matrix} \right] \left[ \begin{matrix} 2 & -1 \\ -1 & 2 \end{matrix} \right] \left[ \begin{matrix} x \\ y \end{matrix} \right] - \left[ \begin{matrix} -3 & 2 \end{matrix} \right] \left[ \begin{matrix} x \\ y \end{matrix} \right] + 4 \\ &=& x^2 + y^2 -xy +3x -2y + 4 . \end{eqnarray*}$$ 



We set:
$$Q =  \left[ \begin{matrix} 2 & -1 \\ -1 & 2 \end{matrix} \right] \qquad  L_1 = \left[ \begin{matrix} -3 \\ 2 \end{matrix} \right].$$





---
# Standard form for Quadratic Progarmming

three decision variables: x1, x2, x3

$$\begin{align*}     \text{minimize: }   - 5 x_2 + 1/2 (x_1^2 + x_2^2 + x_3^2)\\ \text{subject to: }     -4 x_1 - 3 x_2       & >= -8 \\                         2 x_1 +   x_2       & >= 2 \\                         - 2 x_2 + x_3 & >= 0 \end{align*}$$

Could you put this problem into a matrix form?

---
# Quadratic Progarmming in ROI
```{r}
## Simple quadratic program.
## minimize: - 5 x_2 + 1/2 (x_1^2 + x_2^2 + x_3^2)
## subject to: -4 x_1 - 3 x_2       >= -8
##              2 x_1 +   x_2       >=  2
##                    - 2 x_2 + x_3 >=  0

QP <- OP( Q_objective (Q = diag(1, 3), L = c(0, -5, 0)),   # quadratic objective
          # qudratic part: x'Qx,     linear part: Lx
          L_constraint(L = matrix(c(-4,-3,0,2,1,0,0,-2,1), # linear constraint
                                  ncol = 3, byrow = TRUE), # byrow is easy for reading
                       # Linear:    Lx
                       dir = rep(">=", 3), 
                       rhs = c(-8,2,0)) ) # right hand side, vector b
QP
```


---
# Solve

```{r}
sol <- ROI_solve(QP, solver = "quadprog") # choose quadratic solver
sol
```

---
# Extract solutions
```{r}
sol$solution
# sol$objval
sol$status
sol$message #not much information
```

---
# Solutions

```{r}
solution(sol, type = c("primal"))
solution(sol, type = c("dual")) # quadratic solver is not good at it.
```




---
# Portfolio optimization-minimum variance

To minimize the risk, we choose invest in 30 stocks and we need to decide the share for each stock.
Decision variables: x1, x2,..., x30. sum(x, i) = 1.
If xi is allowed to negative, it means we can borrow stock and sell it.


```{r}
## Portfolio optimization - minimum variance
## -----------------------------------------
## get monthly returns of 30 US stocks
suppressMessages(library("ROI"))
##   
data( US30 )  # 180 months for 30 stocks
r <- na.omit( US30 )
head(US30,2)
```
Could you use ROI to solve this questions?

---
# Minimize variance in ROI
```{r}
## objective function to minimize
obj <- Q_objective( 2*cov(r) ) # quadratic objective: x'Mx . M is variance convariance matrix: cov(r) . ## full investment constraint
full_invest <- L_constraint( rep(1, ncol(US30)), "==", 1 ) #  linear constraint for stock shares
#                           (1,1, ..., 1) * (x1,x2,..., x30) ==1
## create optimization problem / long-only
(op <- OP( objective = obj, constraints = full_invest )) 
```
---
# Solution


```{r}
## solve the problem - only works if a QP solver is registered
res <- ROI_solve( op )
res
sol <- solution( res )
names( sol ) <- colnames( US30 )
round( sol[ which(sol > 1/10^6) ], 3 ) # only show large share stock. Not all 30 xi.
```

---
class: inverse, center, middle
name: summary

# Summary

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---
# Summary

- Today,
  - We covered the standard forms for linear programming and quadratic programming problems.
  - We learned how to use ROI to solve linear programming and quadratic programming problems.

---

class: center, middle

# Thanks!

Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).

The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](http://yihui.name/knitr), and [R Markdown](https://rmarkdown.rstudio.com).
